NumLin: Linear Types for Linear Algebra

### 2018-10-28 write-up/paper/implementation.tex
Start code generation subsubsection

Desugaring is conventional, outlined in Figure~\ref{fig:lang_desugar}.  Matrix
expression are translated into BLAS/ LAPACK calls via pattern-matching,
outlined in Figure~\ref{fig:lang_matexp}. Type checking is mostly standard for
a linearly typed language, with the exception of fractional permission
inference.  Because all functions must have their argument types explicitly
annotated, inferring the correct fraction at a call-site is simply a matter of
unification.

@@

\caption{Pattern-matching on matrix expressions.}\label{fig:lang_matexp}
\end{figure}

\subsubsection{Code Generation}

\begin{figure}[tp]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}
A few examples?

>>>>

Desugaring is conventional, outlined in Figure~\ref{fig:lang_desugar}.  Matrix
expression are translated into BLAS/ LAPACK calls via purely syntactic
pattern-matching, outlined in Figure~\ref{fig:lang_matexp}. Type checking is
mostly standard for a linearly typed language, with the exception of fractional
permission inference.  Because all functions must have their argument types
explicitly annotated, inferring the correct fraction at a call-site is simply a
matter of unification.

@@

\caption{Purely syntactic pattern-matching translations of
    matrix expressions.}\label{fig:lang_matexp}
\end{figure}

\subsubsection{Code Generation} is a straightforward mapping from \lang's core
constructs to high-level OCaml ones. We embed \lang's type- and term-
constructors into OCaml as a sanity check on the output
(Figure~\ref{fig:type_grammar}).

\begin{figure}[t]
    \centering
    \begin{minipage}{.3\textwidth}
        \centering
        \begin{grammar}
            <f> ::= `'
            \alt <fc>
            \alt `Z'
            \alt `S' <f>

@@

type 'a bang =
  Many of 'a
  [@@unboxed]
        \end{minted}
    \end{minipage}
    \begin{minipage}{.3\textwidth}
        \begin{align*}
            [\![ f\!c ]\!] &= \texttt{'fc} \\
            [\![ \textbf{Z} ]\!] &= \texttt{z}\\
            [\![ \textbf{S} \, f ]\!] &= [\![ f ]\!]\, \texttt{s}\\
            [\![ \textbf{unit} ]\!] &= \texttt{unit}\\
            [\![ \textbf{bool} ]\!] &= \texttt{bool}\\
            [\![ \textbf{int} ]\!] &= \texttt{int}\\
            [\![ \textbf{elt} ]\!] &= \texttt{float}\\
            [\![ f\, \textbf{arr} ]\!] &= [\![ f ]\!]\, \texttt{arr}\\
            [\![ f\, \textbf{mat} ]\!] &= [\![ f ]\!]\, \texttt{mat}\\
            [\![ \textbf{!} \, t ]\!] &= [\![ t ]\!]\, \texttt{bang}\\
            [\![ \forall f\!c.\, t ]\!] &= [\![ t ]\!]\\
            [\![ t \otimes t' ]\!] &= [\![ t ]\!] \texttt{*} [\![ t' ]\!]\\
            [\![ t \multimap t' ]\!] &= [\![ t ]\!] \rightarrow [\![ t' ]\!]
        \end{align*}
    \end{minipage}
    \caption{\lang's type grammar (left) and its embedding into OCaml
        (right).}\label{fig:type_grammar}
\end{figure}

This is also useful when using \lang\ from within OCaml; for example, we can
use existing tools to inspect the type of the function we are using
(Figure~\ref{fig:build}). It is worth reiterating that only the type- and term-
constructors are translated into OCaml, \lang's precise control over linearity
and aliasing are not brought over.

We actually use this fact to our advantage to clean up the output code by
removing what would otherwise be redundant re-bindings in the output OCaml.
Combined with a code-formatter, the resulting code is not obviously correct and
exactly what a programmer would intend to write by hand, but now with the
guarantees and safety of \lang\ behind it.

\begin{figure}[t]
    \centering
    \includegraphics[width=\textwidth]{impl_build}
    \caption{Using \lang\ functions from OCaml.}\label{fig:build}
\end{figure}

A few examples?

### 2018-10-28 write-up/paper/intro.tex
Start code generation subsubsection

\lang\ is a functional programming language designed to express the APIs of
low-level linear algebra libraries (such as BLAS/LAPACK) safely and explicitly.
It does so by combining linear types, fractional permissions, runtime errors
and recursion into a small, easily understandable, yet expressive set of core
constructs. \lang\ allows a novice to understand and work with complicated
linear algebra library APIs; as well as point out subtle aliasing bugs and
reduce memory usage in existing programs. In addition to this, \lang's
implementation supports several syntactic conveniences as well as a
\emph{usable} integration with real OCaml libraries.

@@

In this paper
\begin{itemize}
    \item we describe \lang, a linearly typed language for linear algebra programs
    \item we illustrate that \lang's design and features are well-suited to its
        intended domain with progressively sophisticated examples
    \item we prove \lang's soundness, using a step-indexed logical relation
    \item we describe a very simple, unification based type-inference algorithm
        for polymorphic fractional permissions (similar to ones used for
        parametric polymorphism), demonstrating an alternative approach to
        dataflow analysis \cite{bierhoff}
    \item we describe an implementation that is both compatible with and usable
        from existing code.
\end{itemize}

>>>>

\lang\ is a functional programming language designed to express the APIs of
low-level linear algebra libraries (such as BLAS/LAPACK) safely and explicitly.
It does so by combining linear types, fractional permissions, runtime errors
and recursion into a small, easily understandable, yet expressive set of core
constructs.

\lang\ allows a novice to understand and work with complicated linear algebra
library APIs, as well as point out subtle aliasing bugs and reduce memory usage
in existing programs. In fact, we were able to use \lang\ to find linearity and
aliasing bugs in a linear algebra program that was \emph{generated} by another
program \emph{specifically designed to translate matrix expressions into an
efficient sequence of calls linear algebra routines}. We were also able to
reduce the number of temporaries used by the same program, using \lang's type
system to guide us.

\lang's implementation supports several syntactic conveniences as well as a
\emph{usable} integration with real OCaml libraries.

@@

In this paper
\begin{itemize}
    \item we describe \lang, a linearly typed language for linear algebra programs
    \item we illustrate that \lang's design and features are well-suited to its
        intended domain with progressively sophisticated examples
    \item we prove \lang's soundness, using a step-indexed logical relation
    \item we describe a very simple, unification based type-inference algorithm
        for polymorphic fractional permissions (similar to ones used for
        parametric polymorphism), demonstrating an alternative approach to
        dataflow analysis \cite{bierhoff}
    \item we describe an implementation that is both compatible with and usable
        from existing code
    \item we show an example of how using \lang\ helped highlight linearity
        and aliasing bugs, and reduce the memory usage of a \emph{generated}
        linear algebra program
    \item we show that using \lang, we can achieve parity with C for linear
        algebra routines, whilst having much better static guarantees about the
        linearity and aliasing behaviour of our programs.
\end{itemize}
